---
title: "STATS 506 Fall 2026 Homework 2"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

github link: <https://github.com/hebe233299-source/stats506-hw2>


```{r}
#| label: setup

library(microbenchmark)

options(digits = 5)
```

# Problem 1

## a.

I first define the four possible directions and check that `n` is a positive
integer and that `p` is between 0 and 1. The function then simulates the walk
step by step and returns the final coordinates.

```{r}
#| label: walk-setup

walk_directions <- rbind(
  right = c(1, 0),
  left = c(-1, 0),
  up = c(0, 1),
  down = c(0, -1)
)

validate_walk_inputs <- function(n, p) {
  valid_n <- is.numeric(n) &&
    length(n) == 1L &&
    !is.na(n) &&
    is.finite(n) &&
    n >= 1 &&
    n == floor(n) &&
    n <= .Machine$integer.max

  if (!valid_n) {
    stop("`n` must be one finite, positive whole number.", call. = FALSE)
  }

  valid_p <- is.numeric(p) &&
    length(p) == 1L &&
    !is.na(p) &&
    is.finite(p) &&
    p >= 0 &&
    p <= 1

  if (!valid_p) {
    stop("`p` must be one finite number between 0 and 1.", call. = FALSE)
  }

  list(n = as.integer(n), p = as.numeric(p))
}
momentum_walk <- function(n, p) {
  inputs <- validate_walk_inputs(n, p)
  n <- inputs$n
  p <- inputs$p

  direction_indices <- integer(n)
  direction_indices[1] <- sample.int(4L, size = 1L)

  if (n > 1L) {
    for (step in 2:n) {
      keep_direction <- runif(1L) < p

      if (keep_direction) {
        direction_indices[step] <- direction_indices[step - 1L]
      } else {
        direction_indices[step] <- sample.int(4L, size = 1L)
      }
    }
  }

  final_position <- colSums(
    walk_directions[direction_indices, , drop = FALSE]
  )

  setNames(as.numeric(final_position), c("x", "y"))
}
```

## b.

```{r}
#| label: walk-functions

momentum_walk1 <- function(n, p) {
  inputs <- validate_walk_inputs(n, p)
  n <- inputs$n
  p <- inputs$p

  direction_indices <- integer(n)
  direction_indices[1] <- sample.int(4L, size = 1L)

  if (n > 1L) {
    for (step in 2:n) {
      keep_direction <- runif(1L) < p

      if (keep_direction) {
        direction_indices[step] <- direction_indices[step - 1L]
      } else {
        direction_indices[step] <- sample.int(4L, size = 1L)
      }
    }
  }

  final_position <- colSums(
    walk_directions[direction_indices, , drop = FALSE]
  )
  setNames(as.numeric(final_position), c("x", "y"))
}

momentum_walk2 <- function(n, p) {
  inputs <- validate_walk_inputs(n, p)
  n <- inputs$n
  p <- inputs$p

  state <- new.env(parent = emptyenv())
  first_direction <- sample.int(4L, size = 1L)
  state$direction <- first_direction

  later_directions <- unlist(
    lapply(seq_len(n - 1L), function(step) {
      if (runif(1L) >= p) {
        state$direction <- sample.int(4L, size = 1L)
      }
      state$direction
    }),
    use.names = FALSE
  )

  direction_indices <- c(first_direction, later_directions)
  final_position <- colSums(
    walk_directions[direction_indices, , drop = FALSE]
  )
  setNames(as.numeric(final_position), c("x", "y"))
}
```

For example, invalid values of n and p return the following errors.

```{r}
#| label: validation-evidence

validation_evidence <- data.frame(
  invalid_input = c("n = 0", "p = 1.2"),
  error_message = c(
    tryCatch(
      momentum_walk1(0, 0.6),
      error = function(error) conditionMessage(error)
    ),
    tryCatch(
      momentum_walk2(10, 1.2),
      error = function(error) conditionMessage(error)
    )
  )
)

knitr::kable(validation_evidence)
```

```{r}
#| label: walk-demonstration

set.seed(50601)
momentum_walk1(10, 0.6)
momentum_walk2(10, 0.6)
momentum_walk1(1000, 0.6)
momentum_walk2(1000, 0.6)
```

momentum_walk1() updates the direction step by step with a for loop.
momentum_walk2() uses lapply() while storing the previous direction in state, so the steps are still generated sequentially. When a new direction is drawn, all four directions are possible, including the previous one. Both functions return the final (x, y) position.

## c.

```{r}
#| label: reproducibility

reproducibility_seed <- 506

set.seed(reproducibility_seed)
walk_10_first <- momentum_walk1(10, 0.6)
set.seed(reproducibility_seed)
walk_10_second <- momentum_walk1(10, 0.6)

set.seed(reproducibility_seed)
walk_1000_first <- momentum_walk1(1000, 0.6)
set.seed(reproducibility_seed)
walk_1000_second <- momentum_walk1(1000, 0.6)

reproducibility_results <- data.frame(
  n = c(10L, 1000L),
  first_x = c(walk_10_first["x"], walk_1000_first["x"]),
  first_y = c(walk_10_first["y"], walk_1000_first["y"]),
  second_x = c(walk_10_second["x"], walk_1000_second["x"]),
  second_y = c(walk_10_second["y"], walk_1000_second["y"]),
  identical = c(
    identical(walk_10_first, walk_10_second),
    identical(walk_1000_first, walk_1000_second)
  ),
  row.names = NULL
)

knitr::kable(reproducibility_results)
```

`set.seed(506)` places R's pseudo-random-number generator in a known state.
Resetting that state before the same sequence of random computations produces
the same sequence of draws, so both repeated simulations have identical final
coordinates, as shown in the last column.

## d.

```{r}
#| label: single-walk-comparison

set.seed(50602)
momentum_values <- c(0, 0.5, 0.9)

final_coordinates <- t(
  vapply(
    momentum_values,
    function(momentum) momentum_walk1(1000, momentum),
    numeric(2L)
  )
)

single_walk_results <- data.frame(
  p = momentum_values,
  x = final_coordinates[, "x"],
  y = final_coordinates[, "y"]
)
single_walk_results$final_distance <- sqrt(
  single_walk_results$x^2 + single_walk_results$y^2
)

knitr::kable(single_walk_results, digits = 3)

distance_changes <- diff(single_walk_results$final_distance)
single_walk_pattern <- if (all(distance_changes > 0)) {
  "increased"
} else if (all(distance_changes < 0)) {
  "decreased"
} else {
  "did not change monotonically"
}
```

The final distances for $p=0$, $0.5$, and $0.9$ are, respectively,
`r paste(format(round(single_walk_results$final_distance, 3), nsmall = 3), collapse = ", ")`.
In these particular simulations, distance `r single_walk_pattern` as momentum
increased. Because this comparison uses only one random walk at each value of
`p`, it is not evidence about typical behavior; the Monte Carlo analysis below
is more informative.

## e.

```{r}
#| label: monte-carlo-walks

set.seed(50603)
monte_carlo_p <- c(0, 0.25, 0.5, 0.75, 0.9)
walks_per_p <- 1000L
steps_per_walk <- 1000L

mean_distance <- vapply(
  monte_carlo_p,
  function(momentum) {
    distances <- replicate(walks_per_p, {
      final_position <- momentum_walk1(steps_per_walk, momentum)
      sqrt(sum(final_position^2))
    })
    mean(distances)
  },
  numeric(1L)
)

monte_carlo_results <- data.frame(
  p = monte_carlo_p,
  mean_distance = mean_distance
)

knitr::kable(monte_carlo_results, digits = 3)

```

The mean final distance increases as `p` increases. It rises from
`r round(mean_distance[1], 3)` at $p=0$ to
`r round(mean_distance[length(mean_distance)], 3)` at $p=0.9$.
This suggests that stronger momentum tends to produce a larger final displacement.

## f.

```{r}
#| label: benchmark-walks

set.seed(50604)
benchmark_1000 <- microbenchmark(
  loop = momentum_walk1(1000, 0.6),
  apply = momentum_walk2(1000, 0.6),
  times = 30L
)
benchmark_100000 <- microbenchmark(
  loop = momentum_walk1(100000, 0.6),
  apply = momentum_walk2(100000, 0.6),
  times = 5L
)

benchmark_1000
benchmark_100000

summarize_benchmark <- function(benchmark_result, n) {
  medians <- aggregate(
    time ~ expr,
    data = as.data.frame(benchmark_result),
    FUN = median
  )
  data.frame(
    n = n,
    implementation = as.character(medians$expr),
    median_milliseconds = medians$time / 1e6,
    row.names = NULL
  )
}

benchmark_summary <- rbind(
  summarize_benchmark(benchmark_1000, 1000L),
  summarize_benchmark(benchmark_100000, 100000L)
)
knitr::kable(benchmark_summary, digits = 3)

small_summary <- subset(benchmark_summary, n == 1000L)
large_summary <- subset(benchmark_summary, n == 100000L)
faster_small <- small_summary$implementation[
  which.min(small_summary$median_milliseconds)
]
faster_large <- large_summary$implementation[
  which.min(large_summary$median_milliseconds)
]
small_gap <- diff(range(small_summary$median_milliseconds))
large_gap <- diff(range(large_summary$median_milliseconds))
gap_comparison <- if (large_gap > small_gap) "larger" else "not larger"
```

The median benchmark shows that the `r faster_small` implementation is faster
for $n=1{,}000$ and the `r faster_large` implementation is faster for
$n=100{,}000$. The absolute median-time difference is
`r round(small_gap, 3)` ms for the smaller input and
`r round(large_gap, 3)` ms for the larger input, so the observed difference is
`r gap_comparison` at $n=100{,}000$.

# Problem 2

```{r}
#| label: bus-simulation

set.seed(50605)
number_of_days <- 10000L

early_morning <- rowSums(
  matrix(rpois(number_of_days * 3L, lambda = 4), nrow = number_of_days)
)
morning_peak <- rnorm(number_of_days, mean = 45, sd = 4)
midday <- rowSums(
  matrix(rpois(number_of_days * 6L, lambda = 15), nrow = number_of_days)
)
afternoon_peak <- rnorm(number_of_days, mean = 55, sd = 5)
evening <- rowSums(
  matrix(rpois(number_of_days * 6L, lambda = 9), nrow = number_of_days)
)

daily_passenger_total <- early_morning + morning_peak + midday +
  afternoon_peak + evening

bus_results <- data.frame(
  simulated_days = number_of_days,
  estimated_mean_passengers = mean(daily_passenger_total),
  monte_carlo_standard_error = sd(daily_passenger_total) /
    sqrt(number_of_days)
)

knitr::kable(bus_results, digits = 3)
```

Using 10,000 simulated days and no loops, the estimated expected daily total is
`r round(bus_results$estimated_mean_passengers, 3)` passengers. The Monte Carlo
standard error is `r round(bus_results$monte_carlo_standard_error, 3)`. The
estimate is very close to the theoretical mean of
$3(4)+45+6(15)+55+6(9)=256$ passengers.

# Problem 3


```{r}
#| label: youtube-data

youtube <- read.csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-03-02/youtube.csv"
)

names(youtube)

data_types <- data.frame(
  variable = names(youtube),
  type = vapply(youtube, function(variable) class(variable)[1], character(1L)),
  row.names = NULL
)
knitr::kable(data_types)

binary_characteristics <- c(
  "funny",
  "show_product_quickly",
  "patriotic",
  "celebrity",
  "danger",
  "animals",
  "use_sex"
)

binary_evidence <- data.frame(
  characteristic = binary_characteristics,
  false_count = vapply(
    youtube[binary_characteristics],
    function(variable) sum(!variable, na.rm = TRUE),
    integer(1L)
  ),
  true_count = vapply(
    youtube[binary_characteristics],
    function(variable) sum(variable, na.rm = TRUE),
    integer(1L)
  ),
  missing = vapply(
    youtube[binary_characteristics],
    function(variable) sum(is.na(variable)),
    integer(1L)
  ),
  row.names = NULL
)
knitr::kable(binary_evidence)
```

## a.

```{r}
#| label: deidentify-youtube

identifying_variables <- c(
  "brand",
  "superbowl_ads_dot_com_url",
  "youtube_url",
  "id",
  "etag",
  "published_at",
  "title",
  "description",
  "thumbnail",
  "channel_title"
)

youtube_deid <- youtube[
  , setdiff(names(youtube), identifying_variables),
  drop = FALSE
]

youtube_deid[binary_characteristics] <- lapply(
  youtube_deid[binary_characteristics],
  as.integer
)

deidentification_results <- data.frame(
  rows = nrow(youtube_deid),
  columns = ncol(youtube_deid),
  removed_columns = paste(identifying_variables, collapse = ", ")
)
knitr::kable(deidentification_results)
```

The de-identified data have `r nrow(youtube_deid)` rows and
`r ncol(youtube_deid)` columns. I removed the brand, both commercial/video
URLs, the YouTube ID and API tag, publication timestamp, title, description,
thumbnail URL, and channel title because those fields could identify a
particular commercial. The seven logical characteristics were retained and
encoded as 0/1 variables; `year` and all five engagement variables were also
retained.

## b.

```{r}
#| label: outcome-diagnostics

engagement_variables <- c(
  "view_count",
  "like_count",
  "dislike_count",
  "favorite_count",
  "comment_count"
)

summarize_engagement <- function(variable_name) {
  values <- youtube_deid[[variable_name]]
  observed <- values[!is.na(values)]

  data.frame(
    variable = variable_name,
    observed = length(observed),
    missing = sum(is.na(values)),
    zeros = sum(observed == 0),
    proportion_zero = mean(observed == 0),
    minimum = min(observed),
    median = median(observed),
    mean = mean(observed),
    percentile_95 = unname(quantile(observed, 0.95)),
    maximum = max(observed),
    row.names = NULL
  )
}

engagement_summary <- do.call(
  rbind,
  lapply(engagement_variables, summarize_engagement)
)
knitr::kable(engagement_summary, digits = 3)
```

```{r}
#| label: outcome-histograms
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Empirical distributions. Nonconstant outcomes are shown on the log1p scale; favorite count is shown on its original scale."

old_graphics_parameters <- par(mfrow = c(2, 3), mar = c(4, 4, 3, 1))
invisible(
  lapply(engagement_variables, function(variable_name) {
    values <- youtube_deid[[variable_name]]
    is_constant <- length(unique(values[!is.na(values)])) == 1L
    plotted_values <- if (is_constant) values else log1p(values)
    axis_label <- if (is_constant) {
      variable_name
    } else {
      paste0("log1p(", variable_name, ")")
    }

    hist(
      plotted_values,
      breaks = 20L,
      main = gsub("_", " ", variable_name),
      xlab = axis_label,
      col = "gray80",
      border = "white"
    )
  })
)
plot.new()
par(old_graphics_parameters)
```

```{r}
#| label: outcome-decisions

outcome_decisions <- data.frame(
  original_variable = engagement_variables,
  category = c(
    "2. Transform before linear regression",
    "2. Transform before linear regression",
    "2. Transform before linear regression",
    "3. Not appropriate for linear regression",
    "2. Transform before linear regression"
  ),
  transformation = c(
    "log1p(view_count)",
    "log1p(like_count)",
    "log1p(dislike_count)",
    "None",
    "log1p(comment_count)"
  ),
  justification = c(
    "Nonnegative and strongly right-skewed with extreme values.",
    "Nonnegative and strongly right-skewed with extreme values.",
    "Right-skewed count with zeros; log1p retains zeros.",
    "All observed values are zero, so there is no outcome variation.",
    "Right-skewed count with zeros and missing values; log1p retains zeros."
  )
)
knitr::kable(outcome_decisions)

usable_original_outcomes <- c(
  "view_count",
  "like_count",
  "dislike_count",
  "comment_count"
)
model_outcomes <- paste0("log_", usable_original_outcomes)

youtube_deid[model_outcomes] <- lapply(
  youtube_deid[usable_original_outcomes],
  log1p
)

outcome_labels <- setNames(
  paste0("log1p(", usable_original_outcomes, ")"),
  model_outcomes
)
```

The summaries and histograms show extreme right skew for views, likes,
dislikes, and comments, so I use `log1p()` for those four outcomes; this
transformation is defined at zero. `favorite_count` is zero for every observed
commercial and therefore cannot serve as a regression outcome. Missing
comments are left missing and are handled by complete-case fitting in each
model.

## c.

```{r}
#| label: regression-models

fit_characteristic_model <- function(outcome, characteristic) {
  model_formula <- reformulate(
    c(characteristic, "year"),
    response = outcome
  )
  fitted_model <- lm(model_formula, data = youtube_deid)
  coefficient_table <- summary(fitted_model)$coefficients

  data.frame(
    outcome = outcome_labels[[outcome]],
    characteristic = characteristic,
    estimate = coefficient_table[characteristic, "Estimate"],
    p_value = coefficient_table[characteristic, "Pr(>|t|)"],
    row.names = NULL
  )
}

model_results <- do.call(
  rbind,
  lapply(model_outcomes, function(outcome) {
    do.call(
      rbind,
      lapply(
        binary_characteristics,
        function(characteristic) {
          fit_characteristic_model(outcome, characteristic)
        }
      )
    )
  })
)

display_model_results <- transform(
  model_results,
  estimate = sprintf("%.4f", estimate),
  p_value = format.pval(p_value, digits = 4, eps = 0.0001)
)
knitr::kable(display_model_results)
```

At $\alpha=0.05$, the following table gives every statistically significant
association and its direction.

```{r}
#| label: significant-regressions

significant_results <- subset(model_results, p_value < 0.05)
significant_results$direction <- ifelse(
  significant_results$estimate > 0,
  "positive",
  "negative"
)

display_significant_results <- transform(
  significant_results,
  estimate = sprintf("%.4f", estimate),
  p_value = format.pval(p_value, digits = 4, eps = 0.0001)
)
knitr::kable(display_significant_results)
```

At $\alpha=0.05$, none of the seven commercial characteristics is statistically significant for any of the four engagement outcomes.

```{r}
#| label: largest-view-association

view_outcome <- "log_view_count"
view_results <- subset(
  model_results,
  outcome == outcome_labels[[view_outcome]]
)
positive_view_results <- subset(view_results, estimate > 0)
largest_positive_view <- positive_view_results[
  which.max(positive_view_results$estimate),
  ,
  drop = FALSE
]

knitr::kable(
  transform(
    largest_positive_view,
    estimate = sprintf("%.4f", estimate),
    p_value = format.pval(p_value, digits = 4, eps = 0.0001)
  )
)
```

The largest positive estimated association with the view outcome is for
`r largest_positive_view$characteristic`, with coefficient
`r round(largest_positive_view$estimate, 4)`. Because the response is
`log1p(view_count)`, this is a difference on the transformed scale after
adjusting for continuous `year`, not a raw increase in the number of views.

## d.

```{r}
#| label: matrix-regression

selected_view_result <- view_results[
  which.min(view_results$p_value),
  ,
  drop = FALSE
]
selected_characteristic <- selected_view_result$characteristic

selection_evidence <- data.frame(
  view_outcome = outcome_labels[[view_outcome]],
  selected_characteristic = selected_characteristic,
  p_value = selected_view_result$p_value
)
knitr::kable(
  transform(
    selection_evidence,
    p_value = format.pval(p_value, digits = 5, eps = 0.00001)
  )
)

matrix_data <- youtube_deid[
  , c(view_outcome, selected_characteristic, "year"),
  drop = FALSE
]
matrix_data <- matrix_data[complete.cases(matrix_data), , drop = FALSE]

design_matrix <- cbind(
  "(Intercept)" = 1,
  matrix_data[[selected_characteristic]],
  matrix_data$year
)
colnames(design_matrix) <- c(
  "(Intercept)",
  selected_characteristic,
  "year"
)
response_vector <- matrix_data[[view_outcome]]

manual_coefficients <- solve(
  t(design_matrix) %*% design_matrix
) %*% t(design_matrix) %*% response_vector

manual_coefficient_table <- data.frame(
  term = rownames(manual_coefficients),
  manual_estimate = as.numeric(manual_coefficients),
  row.names = NULL
)
knitr::kable(manual_coefficient_table, digits = 8)
```

The preceding calculation constructs the intercept, selected characteristic,
and continuous-year columns explicitly and evaluates
$(X^T X)^{-1}X^T y$ without calling `lm()`.

```{r}
#| label: matrix-regression-verification

verification_formula <- reformulate(
  c(selected_characteristic, "year"),
  response = view_outcome
)
verification_model <- lm(verification_formula, data = matrix_data)
lm_coefficients <- coef(verification_model)

coefficient_comparison <- data.frame(
  term = names(lm_coefficients),
  manual = as.numeric(manual_coefficients),
  lm = as.numeric(lm_coefficients),
  absolute_difference = abs(
    as.numeric(manual_coefficients) - as.numeric(lm_coefficients)
  ),
  row.names = NULL
)

knitr::kable(coefficient_comparison, digits = 10)

coefficient_agreement <- isTRUE(
  all.equal(
    as.numeric(manual_coefficients),
    as.numeric(lm_coefficients),
    tolerance = 1e-8
  )
)
maximum_difference <- max(coefficient_comparison$absolute_difference)
```

The matrix calculation and `lm()` coefficients agree up to numerical
precision: `all.equal()` is `r coefficient_agreement`, and the maximum absolute
difference is `r format(maximum_difference, scientific = TRUE, digits = 3)`.
